Notes and other text

### 2023-03-31 doc/manuscript/translocation.qmd
Additional edits to Methods and figure legends

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Bree comment: Nuances of word choice -->.
Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
To date, the use of resistant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
<!--# Other examples in addition to Joseph and Knapp (2018)? In other taxa besides amphibians? -->

@@

In the current study, we combine results from novel <!--# better adjective than "novel". Mark here: Reads OK without an adjective --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"? Mark here:  This transition reads fine to me. -->.
We then reintroduce these resistant frogs to 24 sites and describe the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.

@@

To assess whether frog populations have displayed an evolutionary response following Bd invasion, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
Our goals were to identify putative genes and regions of the genome under selection and assess whether these genes could be related to increased frog resistance to Bd.
Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("persistent"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
<!--# Modify text/remove text here/elsewhere to avoid repetition of this. --> Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.

To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive vs. persistent) and individual variants.
Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types <!--# Reverse the order of these two objectives? Mark here:  I am actually liking the ordering of the objectives now. -->.
Finally, we assessed whether putative genes or genomics regions that showed strong evidence of selection were associated with biological functions consistent with possible resistance mechanisms.
<!--# Could we move the following sentence to the methods? --> We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01) , or a threshold for the W-statistic for the splined window analysis (4 standard deviations from the mean).

Regarding our first analysis, individual frogs clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
The clustering reflected the strong signature of spatial proximity on genetic structure, as has been previously documented with this species <!--# Mark wrote this...confirm validity -->.
We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01).
The liberal set of variants contained 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs (@fig-selectionresults C, D: SI).
The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, Table 2: SI <!--# add this table -->).
For these SNPs, frequency differences between the persistent and recovering groups ranged from 41% to 86%.
Most SNPs in the stringent set showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors.
However, the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C), suggestive of strong selection at this locus associated with exposure of frogs to Bd.
No biological functions, molecular functions, or cellular processes (i.e., GO terms) were significantly overrepresented in the set of 29 annotated outlier genes.

Our second analysis used a splined window to identify regions of the genome under selection.
We found regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and persistent populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$) <!--# Change to delta pi instead of pi_diff? -->.
We identified 58 outlier windows for F~ST~ and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs. <!--# SNPs from the liberal or strict set? Also "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? -->.
A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (SI Dataset 2).
The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$), occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.  Readers not familiar with amphibian-Bd systems might miss the importance of this result --> Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.

Together, our genomics results show that persistent populations have strong signatures of selection on multiple genes and regions of the genome.
While the functions of all genes and regions showing signatures of selection were not clearly interpretable biologically, regions of the genome associated with resistant phenotypes in other amphibian systems such as the extended MHC Class I, showed strong evidence of selection in persistent MYL frog populations.

<!--# Need summary of importance of genetic results, also need better integration of different selection  results.  Mark W: I took a crack at this to test my naive understanding of the genetic results. Figured I could be a good test case of how a genetically naive reader would summarize the key results, to see if our genomic message is coming across clearly. -->

<!--# Based on Fig. 2, there are numerous outlier windows. How were the specific windows mentioned above chosen? Seems important to mention to avoid the perception that mentioned patterns were chosen haphazardly or cherry-picked based on unknown criteria. -->

@@

Results from frog translocations indicated that adult survival was often relatively high and that nearly all populations showed evidence of successful reproduction (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
<!--# May need to add some text early in this section to clarify why so far in this manuscript we have focursed on yearly survival probability of translocated adults, but the model focuses more on yearly survival probability of naturally recruited adults (e.g., sensitivity analysis, etc.). It is my understanding that you are generally assuming that these two probabilities are equivalent and interchangable, but additional clarification would be helpful. Might also require some clarification in Table 2.-->

Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd), our model <!--# Is this a different model, now with demographic stochasiticity and environmental variability in juvenile recruitment, or the same model used to estimate lamda in the previous paragraph? Clarify.  -->predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
<!--# The wording in the previous two sentences seems overly cautious in describing the extinction probability (i.e., "less then 1", "some probability of persistence". In fact, as the next sentence makes clear, for reasonable probabilities of recruitment, the extinction probabilities are much less than 1. Consider tweaking wording to clarify this. I'd argue for a relatively strong initial statement that we then fine tune only slightly, instead of a weak statement that we then strengthen a lot. Initial impressions can be hard to change. --> For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5 <!--# Based on Fig. 5, looks like extinction propability is < 0.5 for all six populations only for omega > ~0.22. Reword as necessary. -->, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
Overall, our model demonstrates that given observed yearly adult survival probabilities of resistant frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
Moreover, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013.
Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.

@@

Disease-induced declines are a central conservation challenge for wildlife and identifying biologically grounded strategies for assisted recovery is a critical need across systems.
One such strategy requires leveraging disease-induced genetic changes in host populations and so-called evolutionary rescue to recover a species at the landscape scale.
While this strategy has been previously proposed, it has never been successfully implemented.
Here, we provide strong evidence that selection for resistance and evolutionary rescue has led to the recovery of some populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
Leveraging individuals from these resistant and rescued populations, we use 15 years of translocation effort and population viability modeling to demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how natural evolutionary rescue can be leveraged to successfully recover a species in the presence of disease.

While conclusively demonstrating that evolutionary rescue is driving population recoveries is challenging.
we provide four lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance and recoveries in MYL frog systems.
First, in host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain long-lasting immunity to Bd and are susceptible to re-infection and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) in the MYL frog system.
Thus, MYL frog-Bd systems have the epidemiological characteristics necessary for evolutionary rescue to occur.
Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering population.
These differences are consistent with strong signatures of selection and regions of the genome under selection are associated with known mechanisms of resistance to Bd infection.
Third, our translocations allowed us to conclude that observed resistance and recovery of resistant populations is not strictly due to local environmental conditions (though environment does play a role).
Changes in population growth rate due to evolved resistance extend beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frog populations in new habitats in the presence of Bd.
Finally, evolved resistance has a quantifiable phenotype.
In persistent populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
While other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), our study is unique in that we show that evolved resistance is associated with population recovery in both natal and non-natal habitats, providing an actionable strategy for promoting the long-term viability of a species.

Despite the results we present here, recovery of Bd-impacted amphibians worldwide is expected to be challenging at best.
Hundreds of declining species, but mostly failed attempts at recovery.
Failures may be due in part to a general inability by amphibians to mount an effective immune response against Bd infection.
In contrast, in our study MYL frogs (a species highly susceptible to chytridiomycosis) showed relatively high frog survival despite ongoing Bd infection and low probability of extinction for many reestablished populations.

@@

To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet R package [@jombart2008].
Next, to detect regions of the genome under selection between naive and recovering populations, we used a multivariate linear mixed model to evaluate individual variants and a splined window analysis to evaluate larger genomic regions.
 <!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.  For example, for use genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection. The first approach A, did X, Y, and Z but had C limitations.  The second approach B, did X, Y, and Z and D limitations."-->.
GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
For this analysis we used a stringent data filter to include only SNPs with \<5% missing data (missing for no more than 2 individuals).
GEMMA then calculated the likelihood ratio statistic <!--# What is the null with this test (or what are the two models being compared with the LRT)?  Probably obvious for genomics folks, but not totally clear to me.  I am guessing we are testing whether the `population` coefficient (naive vs. persistent) is significant for describing variability in a SNP? --> for the set of 148,307 high quality SNPs across 127 contigs.
We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
<!--# Could we say something like: "Outliers indicated that a SNP was differently represented in naive vs. persistent populations" or whatever is the correct language." --> Results were visualized using a Manhattan plot and qqplot.

To detect regions of the genome under selection <!--# Improve wording.  Also, can we contrast this with the previous approach? What is this analysis doing that the other one didn't? -->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
First, we calculated per-site F~ST~ between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations individuals using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
We concatenated the values for F~ST~ and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.

We then used the package WINGEN R package [@bishop2023] <!--# Correct citation? --> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015] .
This method also calculates a W-statistic allowing for outlier identification.
We identified outliers as those with a W-statistic \> overall mean + 4 \* standard deviation for F~ST~ or overall mean +/- 4 \* standard deviation for $\pi_{diff}$.
Outlier regions were then identified as those that were outliers in both analyses, meaning that they showed 1) high differentiation between naive and recovering populations, and 2) differential patterns of nucleotide diversity in the same region.
We then extracted gene transcripts mapped within each region and retrieved annotation for that region using blast.

@@

To identify important predictors of frog survival following translocation, we developed multilevel models in a Bayesian framework.
We performed all analyses using the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022] .
We used default, weakly informative priors.
Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013, @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016].
Both approaches utilized the loo package [@vehtari2022].

>>>>

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Bree comment: Nuances of word choice -->.
In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021]<!--# The previous sentence might fit better in the Discussion when we take a deeper dive into evolutionary rescue -->.
Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
To date, the use of resistant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
<!--# Other examples in addition to Joseph and Knapp (2018)? In other taxa besides amphibians? -->

@@

In the current study, we combine results from novel <!--# better adjective than "novel" --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"? -->.
We then reintroduced these resistant frogs to 24 sites and described the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.

@@

To identify putative genes under selection and potentially related to increased frog resistance to Bd, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees (@fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
These "recovering" populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
<!--# Modify text/remove text here/elsewhere to avoid repetition of this. --> Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
To detect regions of the genome under selection, we compared exome sequences of frogs from naive and recovering populations using (1) a multivariate linear mixed model to evaluate associations between population type and individual variants, and (2) a splined window analysis to evaluate larger genomic regions showing differentiation between population types.
For single SNPs, we identified outlier variants using a Bonferroni-corrected significance level (liberal = 0.05, stringent = 0.01), and for the splined window analysis we used a W-statistic of 4 standard deviations from the mean.
<!--# This is different from the definition in the Methods: 4 * standard deviation -->

The samples clustered into three separate groups in principal component space, each containing at least one population from both the naive and recovering groups (@fig-selectionresults B: SI).
The liberal set of variants contained 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs (@fig-selectionresults C, D: SI).
One gene in this group does not have an associated annotation.
No biological functions, molecular functions, or cellular processes <!--# Are these the "GO terms" referred to in the next paragraph? If so, add here paranthetically. --> were significantly overrepresented in the set of 29 annotated outlier genes.
The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, SI Dataset 1).
For these SNPs, frequency differences between the naive and recovering populations ranged from 41% to 86%.
Most SNPs in the stringent set showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors.
However, the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C), suggestive of strong selection at this locus associated with exposure of frogs to Bd.

In the splined window analysis, we identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both *F~ST~* and $\pi_{diff}$ ("shared" regions) and 2 of the 9 shared regions also contained the outlier SNPs described above <!--# SNPs from the liberal or strict set? Also "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? -->.
A total of 33 annotated genes were found in the 9 shared regions and no GO terms were overrepresented within this set of genes (SI Dataset 2).
The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had higher $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
<!--# Based on Fig. 2, there are numerous outlier windows. How were the specific windows mentioned above chosen? Seems important to mention to avoid the perception that mentioned patterns were chosen haphazardly or cherry-picked based on unknown criteria. -->

@@

Results from frog translocations indicated that adult survival was often relatively high and that nearly all populations showed evidence of successful reproduction (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
<!--# May need to add some text early in this section to clarify why so far in this manuscript we have focused on yearly survival probability of translocated adults, but the model focuses more on yearly survival probability of naturally recruited adults (e.g., in the sensitivity analysis, etc.). It is my understanding that you are generally assuming that these two probabilities are equivalent and interchangable, but additional clarification would be helpful. Might also require some clarification in Table 2. You describe this in more detail in the Methods, but it seems that some additional clarification might be helpful here. -->

Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd), our model <!--# Is this a different model, now with demographic stochasiticity and environmental variability in juvenile recruitment, or the same model used to estimate lamda in the previous paragraph? Clarify. I know from the Methods that you developed different versions of the model, some deterministic and some stochastic, but some clarification might be useful here too. -->predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B).
Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction.
<!--# The wording in the previous two sentences seems overly cautious in describing the extinction probability (i.e., "less than 1", "some probability of persistence". In fact, as the next sentence makes clear, for reasonable probabilities of recruitment, the extinction probabilities are substantially less than 1. Consider tweaking wording to clarify this. I'd argue for a relatively strong initial statement that we then fine tune only slightly, instead of a weak statement that we then strengthen a lot. Initial impressions can be hard to change. --> For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5 <!--# Based on Fig. 5, looks like extinction propability is < 0.5 for all six populations only for omega > ~0.22. Reword as necessary. -->, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).
Overall, our model demonstrates that given observed yearly adult survival probabilities of resistant frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
Moreover, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013.
Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.
<!--# Mark here: Just figured I would point this out. This is trending in the right direction, but there isn't strong statistical evidence that these two proportions differ.  I tried to phrase this carefully. -->

@@

1.  Summary of all of the results in a way that integrates the different components/sections (selection, translocations, viability), and makes the pitch for how unique the results are (as per comment from Bree):
    -   Evidence for evolutionary rescue in a conservation-relevant species.
    -   Successful reestablishment of extirpated populations via translocation despite ongoing Bd infection.
    -   Evidence for long-term viability of reestablished populations.

<!-- -->

2.  Evolutionary rescue: We provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance.
    -   Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
    -   Differences in allele frequencies exist between adjacent Bd-naive and recovering populations.
    -   There is a shared selective response across recovering populations.
    -   Changes in population growth rate due to evolution extend beyond the natal environment.
    -   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).

@@

To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet R package [@jombart2008].
Next, to detect regions of the genome under selection between naive and recovering populations, we used a multivariate linear mixed model to evaluate individual variants and a splined window analysis to evaluate larger genomic regions.
The multivariate linear mixed model was implemented in the program GEMMA [@zhou2014].
GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
For this analysis we used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals).
GEMMA then calculated the likelihood ratio statistic for the set of 148,307 high quality SNPs across 127 contigs.
We calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
Results were visualized using a Manhattan plot and qqplot.

We conducted the splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
First, we calculated per-site *F~ST~* between individuals from naïve and recovering populations for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
We concatenated the values for *F~ST~* and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
We then used the package WINGEN R package [@bishop2023] <!--# Correct citation? --> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015] .
This method also calculates a W-statistic allowing for outlier identification.
We identified outliers as those with a W-statistic \> overall mean + 4 \* standard deviation for F~ST~ or overall mean +/- 4 \* standard deviation for $\pi_{diff}$.
Outlier regions were then identified as those that were outliers in both analyses, meaning that they showed 1) high differentiation between naive and recovering populations, and 2) differential patterns of nucleotide diversity in the same region.
We then extracted gene transcripts mapped within each region and retrieved annotation for that region using blast.

@@

To identify important predictors of frog survival following translocation, we developed multilevel models in a Bayesian framework.
We performed all analyses with the rstanarm package [@rstanarm2022] and R Statistical Software [v4.4.4, @rsoftware2022], and used default, weakly informative priors.
Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013, @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016].
Both approaches utilized the loo package [@vehtari2022].
